---
title: "SideStep15_microRNA_from_normal_RNAseq"
author: "ieva"
date: "2023-10-26"
output: word_document
---
This rmd file is for doing analysis of microRNA but not downloading micorRNA 
specific data, rather filtering for microRNA genes
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = '~/rprojects/TCGA-OV-data')
```
Starting libraries
```{r}
library(tidyverse)
library(dendextend)
library(WGCNA)
library(GDCRNATools)
library(edgeR)
library(limma)
library(DESeq2)
library(glmnet)
```
Starting form step3: tcga and gtex data with all descriptions
desccriptor colums:1:2, 183:184, 601:603
```{r}
gtcga_final <- readRDS("gtcga_final.RDS")
```

```{r}
#remove non-miRNA
biotypes <- table(gtcga_final$gene_biotype, useNA = "a")
biotypes <- as.data.frame(biotypes)
biotypes
gtgca_miRNA <- gtcga_final[(gtcga_final$gene_biotype == "miRNA" ), ]
#test if split correctly
dim(gtgca_miRNA) #liko 1573 genu is 55469     
table(gtgca_miRNA$gene_biotype, useNA = "a") 
```
```{r}
#fix names
gtgca_miRNA["external_gene_name.y"][gtgca_miRNA["external_gene_name.y"] == ''] <- NA
any(is.na(gtgca_miRNA$external_gene_name.y))
gtgca_miRNA <- gtgca_miRNA %>% 
  mutate(external_gene_name.y = coalesce(external_gene_name.y, Description))  
any(is.na(gtgca_miRNA$external_gene_name.y))
#add gene names as rownames
rownames(gtgca_miRNA) <- gtgca_miRNA$external_gene_name.y
head(rownames(gtgca_miRNA))
gtgca_miRNA <- gtgca_miRNA[, -c(1:2, 183:184, 601:603)] 
saveRDS(gtgca_miRNA, "gtgca_miRNA.RDS")
```
Is it even possible to do analysis in such way?
what is the mean count of eatch microrna?
```{r}
means <- rowSums(gtgca_miRNA)
mean(means) #on average 709 counts
length(which(means == 0)) #tik 91 miR no expression anywhere
```
## Outlier detection
```{r}
counts_gtcga_miR <- data.matrix(gtgca_miRNA) 
dim(counts_gtcga_miR)#19193   596
# detect outlier genes with gsg
gsg <- goodSamplesGenes(counts_gtcga_miR) #no transpose, wgcna package
summary(gsg) #weather there are outliers
gsg$allOK #fals tells there are outliers

table(gsg$goodGenes) #false is the number of outliers
table(gsg$goodSamples) #no outliers
```
clustering to see outliers
```{r}
htree <- hclust(dist(t(gtgca_miRNA)), method = "average") #can take some time 
png(file="figures/htree_miRNA1.png", height=3000, width=6000) #išsaugijimui didesniu formatu
plot(htree) #dstant samples should be excluded #save portrait 20x66 inch pdf
dev.off()
```
i will not remove anyone for now
now for the clustering bewteen projects
```{r}
dend <- as.dendrogram(htree)
col_aa_red <- ifelse(grepl("GTEX", labels(dend)), "red", "blue")
dend2 <- assign_values_to_leaves_edgePar(dend=dend, value = col_aa_red, edgePar = "col") 
plot(dend2)
```
## Normalize 
GDC RNA tools, with filter, model added (issitraukus GDC RNA tools funkcija)
```{r}
counts <- counts_gtcga_miR
expr = DGEList(counts = counts)
expr = calcNormFactors(expr)

snames = colnames(counts_gtcga_miR);
group = substr(snames, 1, 4); #Sets up level information for samples.

keepALL <- rowSums(cpm(expr) > 1) >= 0.5*ncol(counts)
nGenes <- as.numeric(summary(keepALL)[2]) + 
  as.numeric(summary(keepALL)[3])
nKeep <- summary(keepALL)[3]

cat (paste('Total Number of genes: ', nGenes, '\n', sep=''))
cat (paste('Number of genes for downstream analysis: ', nKeep, 
           '\n', sep=''))

exprALL <- expr[keepALL,,keep.lib.sizes = TRUE]
v_filtered <- voom(exprALL, design= model.matrix(~0 + group), plot = TRUE)$E

saveRDS(v_filtered, "microrna_voom_protein.RDS")
```

see the normalization and how many genes left
```{r}
mRNA_voomt <- t(v_filtered)
htree_norm <- hclust(dist(mRNA_voomt), method = "average") #can take some time
dend3 <- as.dendrogram(htree_norm)
col_aa_red <- ifelse(grepl("GTEX", labels(dend3)), "red", "blue")
dend3 <- assign_values_to_leaves_edgePar(dend=dend3, value = col_aa_red, edgePar = "col") 
plot(dend3)
```
## Train / test split
```{r}
gtcga_miR_counts <- t(v_filtered)
gtcga_miR_counts <- as.data.frame(gtcga_miR_counts)
# Set (random-number-generator) seed so that results are consistent between runs
set.seed(18) #choose favorite number
train_ids <- rbinom(nrow(gtcga_miR_counts), size = 1, prob = 0.8) ==1 #choose persentage
#we split clinical data first (probably should be indicated by y instead of x)
gtex_miR_counts_train = gtcga_miR_counts[train_ids, ] 
gtex_miR_counts_test  = gtcga_miR_counts[!train_ids, ] 

dim(gtex_miR_counts_train) #489 119
dim(gtex_miR_counts_test) #106 119

saveRDS(gtex_miR_counts_train, "train_gtcga_normcounts_miR.RDS")
saveRDS(gtex_miR_counts_test, "test_gtcga_normcounts_miR.RDS")
```
split clinical data as well
```{r}
pheno <- readRDS("tcga_no_weird_pheno_XENA_TCGA.RDS") #full clinical
dim(pheno) #416  69
#the pheno only has the tcga ids, thus the need to filter
snames_train = rownames(gtex_miR_counts_train);
tcga_train_ <- grep("TCGA", snames_train)
tcga_train_ids <- snames_train[tcga_train_]
pheno_train  = pheno[tcga_train_ids, ]  #336 samples
dim(pheno_train)

snames_test = rownames(gtex_miR_counts_test);
tcga_test_ <- grep("TCGA", snames_test)
tcga_test_ids <- snames_test[tcga_test_]
pheno_test = pheno[tcga_test_ids, ]  #79 samples
dim(pheno_test)

#save
saveRDS(pheno_train, "tcga_pheno_train_MIR.RDS")
saveRDS(pheno_test, "tcga_pheno_test_MIR.RDS")
```
## elastic net
prep for 1st model
```{r}
gtex_miR_counts_train <- data.matrix(gtex_miR_counts_train)
group = substr(snames_train, 1, 4); #Sets up level information for samples.
group = as.factor(group)
```
model
```{r}
#clinical feature: gtex or TCGA data
res_gtex_miR = cv.glmnet(
  x = gtex_miR_counts_train,
  y = group,
  alpha = 0.5,
  family = "binomial"
)
res_gtex_miR 
```
abiem atvejais po 39 atrenka
```{r}
plot(res_gtex_miR)
```
Getting genes that contribute for the prediction
```{r}
res_coef_gtex_miR = coef(res_gtex_miR, s="lambda.min") # the "coef" function returns a sparse matrix
# get coefficients with non-zero values
res_coef_gtex_miR = res_coef_gtex_miR[res_coef_gtex_miR[,1] != 0,] 
res_coef_gtex_miR = res_coef_gtex_miR[-1]
res_coef_gtex_miR_names = names(res_coef_gtex_miR) # get names of the (non-zero) variables.
res_coef_gtex_miR_names 
```
What about another lambda?
```{r}
res_coef_gtex_miR2 = coef(res_gtex_miR, s=0.1) # the "coef" function returns a sparse matrix
# get coefficients with non-zero values
res_coef_gtex_miR2 = res_coef_gtex_miR2[res_coef_gtex_miR2[,1] != 0,] 
res_coef_gtex_miR2 = res_coef_gtex_miR2[-1]
res_coef_gtex_miR_names2 = names(res_coef_gtex_miR2) # get names of the (non-zero) variables.
res_coef_gtex_miR_names2 
```
save model
```{r}
saveRDS(res_gtex_miR, "MIR_elastic_net_model_gtcga.RDS")
saveRDS(res_coef_gtex_miR_names, "MIRs_gtcga_elastic.RDS")
```

## coxnet
Prep 1: get gene matrix with desired 39 miRs
```{r}
gtex_miR_elastic <- gtex_miR_counts_train[, colnames(gtex_miR_counts_train) %in% res_coef_gtex_miR_names] 

gtex_miR_elastic <- as.data.frame(t(gtex_miR_elastic))
gtex_miR_elastic <- gtex_miR_elastic %>%  dplyr::select(starts_with("TCGA")) 
dim(gtex_miR_elastic) #39mirs 337
```
Prep 2: clinical data:
```{r}

#clinical data
pheno_train <- readRDS("tcga_pheno_train_MIR.RDS") 
#337 samples, 69 variables

# survival df
clin_df = pheno_train[,
                   c("barcode",
                     "vital_status",
                     "days_to_death",
                     "days_to_last_follow_up",
                     "neoplasmhistologicgrade",
                     "STAGE")]

# create a new boolean variable that has TRUE for dead patients and FALSE for live patients
clin_df$deceased = clin_df$vital_status == "Dead"

# create an "overall survival" variable that is equal to days_to_death
# for dead patients, and to days_to_last_follow_up for patients who
# are still alive
clin_df$overall_survival = ifelse(clin_df$deceased,
                                  clin_df$days_to_death,
                                  clin_df$days_to_last_follow_up)
# remove 1 patient with STAGE I
clin_df[which(clin_df$STAGE == "Stage I"), "STAGE"] = NA
table(clin_df$STAGE, useNA = "a")

saveRDS(clin_df, "pheno_survival_onlyMIRS.RDS")
```
add all miRS to clin_df
```{r}
gtex_Mirs_df<- as.data.frame(t(gtex_miR_elastic))
colnames(clin_df)
gtex_Mirs_df$barcode <- rownames(gtex_Mirs_df)
clin_miRdf_joined <- left_join(clin_df, gtex_Mirs_df, by = "barcode")
#gene names cannot have weird symbols for coxnet
colnames(clin_miRdf_joined) <- gsub("[.]", "_", colnames(clin_miRdf_joined)) 
colnames(clin_miRdf_joined) <- gsub("[-]", "_", colnames(clin_miRdf_joined)) 
```
Prep 3: things needed for coxnet function
```{r}
#COXNET!
clin_miRdf_joined2 <- clin_miRdf_joined %>% drop_na(overall_survival) %>% drop_na(deceased) #lost 2 cases
clin_miRdf_joined2$decesed2 <- as.integer(as.logical(clin_miRdf_joined2$deceased))
time <- clin_miRdf_joined2$overall_survival
status <- clin_miRdf_joined2$decesed2
miR_y <- clin_miRdf_joined2[ ,c(8,7)]
names(miR_y) <- c("time", "status")
miR_y <- as.matrix(miR_y)
head(miR_y)
miR_surv_counts <- clin_miRdf_joined2[, 9:47] #pasirinkti norimas mir
```
cox_fit - visas lambdas ismeta
```{r}
cox_fitx_miRS <- glmnet(miR_surv_counts, miR_y, family="cox")
cox_fitx_miRS
plot(cox_fitx_miRS)
cox_fitx_miRS
```
cv.glmnet - min ir max + paveikslas - takes more time
```{r}
miR_surv_counts <- as.matrix(miR_surv_counts)
set.seed(5)
cvfit_miR <- cv.glmnet(miR_surv_counts, miR_y, family = "cox", type.measure = "C") #takes some time
plot(cvfit_miR)
cvfit_miR #lambda min and max
```
get names for 18 best 1se
```{r}
coef_x_miR <- coef(cvfit_miR, s = 0.024579) # best 18
head(coef_x_miR)
coef_x_miR = coef_x_miR[coef_x_miR[,1] != 0,] 
res_coef_cox_names_MIR = names(coef_x_miR) # get names of the (non-zero) variables.
res_coef_cox_names_MIR #18
write.csv(res_coef_cox_names_MIR, "mirs_coxnet_names.csv")

```


